CASTRO: A New Compressible Astrophysical Solver. IV. Performance Portability and GPUs

### 2019-03-22 gpus/paper.tex
more text

Supercomputer architectures have evolved a lot over the past
decades~\citep{bell:2015}.  Shared memory vector machines with custom
processors dominated early, displaced in the early 1990s by
distributed memory, massively parallel architectures built upon
commodity CPUs where the message passing interface (MPI) became the
dominant library for achieving parallelism.  As core counts increased,
threading because increasingly important, and OpenMP became the
standard for threading in high-performance computing, with hybrid
MPI+OpenMP approaches to parallelism.  In the race to increase
performance while keeping electrical power consumption reasonable,
accelerators were increasingly turned to, include GPUs and many-core
chips like the Intel Phi series.  The simulation codes we use in
astrophysics need to keep pace with these developments, or they will
not be able to run on the next generation of supercomputers.  In
particular, GPUs are dominant now, so existing algorithms need to be
ported to run on GPUs, or new codes need to be written from scratch
targeting the new architecture.

@@

In this paper, we describe our approach to performance portability
with the \castro\ multiphysics simulation code.  Our goal is to have
the same computational kernels run efficiently on CPUs (including
manycore processors) and GPUs.  We describe the work we needed to do
to enable this portability and show performance metrics for a reacting
flow problem in astrophysics.

@@

\castro\ has been used for science investigations of various different
progenitor models of Type Ia
supernova~\cite{malone:2014,katz:2016,polin:2019}, exoplanet
dynamics~\cite{ryu:2018}, core-collapse and population III
supernovae~\cite{chen:2014,dolence:2015,chen:2017}, and X-ray
bursts~\cite{astronum:2018}..

@@

\begin{figure*}[t]
\centering
\includegraphics[width=0.33\textwidth]{gpu_1} 
\includegraphics[width=0.33\textwidth]{gpu_2} 
\includegraphics[width=0.33\textwidth]{gpu_3}
\caption{\label{fig:loops} caption}
\end{figure*}

@@

OpenACC

OpenMP

Discuss frameworks vs.\ doing it outselves

@@

\section{Scaling}

@@

This problem uses constant gravity, so all of the physics can be done on the GPUs.

>>>>

Supercomputer architectures have evolved a lot over the past
decades~\citep{bell:2015}.  Shared memory vector machines with custom
processors dominated early, displaced in the early 1990s by
distributed memory, massively parallel architectures built upon
commodity CPUs where the message passing interface (MPI) became the
dominant library for achieving parallelism.  As core counts increased,
threading because increasingly important, and OpenMP became the
standard for threading in high-performance computing, with hybrid
MPI+OpenMP approaches to parallelism.  In the race to increase
performance while keeping electrical power consumption reasonable,
accelerators were increasingly turned to, include GPUs and many-core
chips like the Intel Phi series.  Throughout this evolution, the simulation codes we use in
astrophysics had to continue adapt to keep pace with the new computing paradigms, or they will
not be able to run on the next generation of supercomputers.  In
particular, GPUs are dominant now, so existing algorithms need to be
ported to run on GPUs, or new codes need to be written from scratch
targeting the new architecture.

@@

In this paper, we describe our approach to performance portability
with the \castro\ multiphysics simulation code.  We choose not to use
a framework like Kokkos but instead we implement the mechanism for
offloading compute kernels to GPUs inside of the AMReX adaptive mesh
refinement library itself.  Our goal is to have the same computational
kernels run efficiently on CPUs (including manycore processors) and
GPUs.  Early GPU results with a version of \castro\ were desribed in
\citet{astronum:2017}.  Here, we describe the work we needed to do to
enable this portability and show performance metrics for a reacting
flow problem in astrophysics.

@@

\castro\ has been used for science investigations of various different
progenitor models of Type Ia
supernova~\citep{malone:2014,katz:2016,polin:2019}, exoplanet
dynamics~\citep{ryu:2018}, core-collapse and population III
supernovae~\citep{chen:2014,dolence:2015,chen:2017}, and X-ray
bursts~\citep{astronum:2018}.

@@

\begin{figure*}[t]
\centering
\includegraphics[width=0.33\textwidth]{gpu_1} 
\includegraphics[width=0.33\textwidth]{gpu_2} 
\includegraphics[width=0.33\textwidth]{gpu_3}
\caption{\label{fig:loops} Examples of distributing the work inside a
  grid patch.  On the left is the pure MPI approach.  An entire patch
  is assigned to a MPI task and the work loop runs over all of the
  cells in the grid, $\mathtt{lo:hi}$.  At the center is the
  tile-based MPI+OpenMP approach.  Again the grid is owned by an MPI
  task, but now it is logically subdivided into files (two shown
  here), and each tile is assigned to an OpenMP thread.  Each thread
  loops only over its tile, where the bounds are now given as
  $\mathtt{lo_{tile}:hi_{tile}}$.  Finally on the right is a GPU
  approach.  As with the others, the grid is owned by an MPI task but
  now each zone is assigned to a different GPU thread on the
  accelerator.  This is achieved by calling the same compute kernels,
  with $\mathtt{lo_{gpu}:hi_{gpu}}$ now corresponding to a single cell.}
\end{figure*}

@@

@@

The reaction update requires solving a system of ordinary differential
equations (ODEs).  Astrophysical reaction networks are notoriously
stiff, so implicit integrators are used (see,
e.g.\ \citealt{timmes:1999}).  \castro\ uses the VODE integrator
\citep{vode} which implements a variable order backward difference
implicit method to advance the system.  This requires the network to
provide a Jacobian as well as a routine to evaluate the righthand side
of the reaction ODE system.  It is these evaluations that dominate the
floating point work.  For integrating the system on GPUs, our approach
is to put the entire ODE integration on GPUs: the evaluation of the
righthand side and Jacobian as well as the driver which subcycles in
time to integrate the system to the desired tolerance.  VODE is
written in old Fortran, with numerous deprecated features that are not
supported in CUDA Fortran (like {\tt go to} statements and {\tt
  common} blocks).  We updated the code to eliminate these unsupported
statements, using more modern Fortran constructs.  The resulting
codebase is called VODE90 and serves as our GPU integrator.  \MarginPar{more here from Don}

\section{Performance}

On summit, we create a resource set with 1 MPI, 1 GPU, and 7 threads,
and there are 6 resource sets per node.

Do we also want to run on Cori?

@@

The {\tt flame\_wave} problem in \castro\ was described in
\cite{astronum:2018}.  This problem explores the lateral propagation
of a helium deflagration across a neutron star in a plane-parallel
atmosphere.  The Castro PPM CTU hydrodynamics solver is used, with a
constant gravitational acceleration, and the problem is solved in the
corotating frame, where the Coriolis forces acts to balance the
lateral spreading in the atmosphere via geostrophic balance.  For the
microphysics, we use the general stellar equation of state of
\cite{timmes_swesty:2000} and the stellar conductivity described in
\cite{Timmes00}.  The science goal of the problem is to understand the
dynamics of a resolved burning front in X-ray bursts.  For this work,
this problem serves as an ideal reacting flow problem to benchmark
GPUs with.  All of the necessary physics: hydrodynamics, equation of
state, thermal conductivities, reactions, hydrodynamic source terms,
and thermal diffusion are run on the GPU.  We look at two different
reaction networks, a 7-isotope alpha network and a 13-isotope alpha
network~\cite{iso7}.
